Finding Maximum $k$-Cliques Faster using Lazy Global Domination

### 2016-04-12 paper/paper.tex
Little wording changes

\Cref{table:clustering} presents results from the smallest 20 clustering graphs
from the 10th DIMACS Implementation
Challenge\footnote{http://www.cc.gatech.edu/dimacs10/archive/clustering.shtml}. Again, from a
maximum clique perspective these would be considered unusually large graphs. However, only five were
unsolvable within an hour (plus a sixth when the domination rule was not used), and over half of the
problems took under two seconds.

@@

On the other hand, for ``wing-nodal'' with $k=4$, the parallel run did much less work than the
sequential run: a speedup of over 100 was obtained from 20 cores. A similar effect occurred with
``cond-mat'' and $k=4$. This is because the work splitting mechanism we used explicitly diversifies
at the top of search first, where branching heuristics are least likely to be correct, which leads
to an initial incumbent being found faster---this is in line with recent observations that tailored
work stealing should be favoured over randomised work stealing for combinatorial search problems
\cite{DBLP:conf/cp/ChuSS09,DBLP:journals/topc/McCreeshP15}.

@@

An Erd\H{o}s-R\'{e}nyi random graph $G(n, p)$ has $n$ vertices, and an edge between each distinct
pair of vertices with probability $p$, chosen independently. Here we investigate the size of a
maximum $k$-clique in such graphs, and the complexity of finding it. In each case, we use an average
over 100 samples for every point.  We do not use the domination rule for these experiments: the
probability of random graphs having dominating vertices is very low.

@@

We have shown that using a maximum clique algorithm to solve the maximum $k$-clique algorithm for a
graph $G$ by considering $G^k$ in place of $G$ is feasible in practice. This is despite $G^k$
potentially being dense even if $G$ is sparse---this ruled out the use of maximum clique algorithms
which are designed for sparse graphs
\cite{DBLP:conf/waw/PattabiramanPGLC13,DBLP:journals/cor/SegundoLP16}, and we were considering
graphs with many more vertices than is typical for dense maximum clique algorithms.

We introduced a new lazy global domination rule. This was sometimes extremely beneficial, leading to
exponential reductions in the search space---without this rule, we would have been unable to solve
eight of the problem instances we considered, and many others would have taken much longer. However,
even with laziness there is still sometimes a cost to pay when this rule does nothing. This rule is
thus harmful (although only by a polynomial factor) for the graphs typically considered for the
maximum clique problem, and we see the benefit of tailoring algorithms to the problem being solved.
We suggest that a similar rule may also be useful for the maximum $k$-club problem.

We were able to use parallelism to close two further instances, although this required a much finer
level of task granularity than usual.  We suspect further progress could be made, for example, by
tailoring the initial vertex ordering based upon what we know about the graphs, or by increasing the
number of recursive calls per second by making the colouring stage cheaper, for example by reusing
colour classes \cite{DBLP:conf/lion/NikolaevBS15}.

>>>>

\Cref{table:clustering} presents results from the smallest 20 clustering graphs
from the 10th DIMACS Implementation
Challenge\footnote{http://www.cc.gatech.edu/dimacs10/archive/clustering.shtml}. Again, from a
maximum clique perspective these would be considered unusually large graphs. However, only five were
unsolvable within an hour (plus a further two when the domination rule was not used), and over half of the
problems took under two seconds.

@@

As well as improving the results on the instances we could not solve sequentially, parallel search
sometimes gave large improvements for the easier instances. For ``wing-nodal'' with $k=4$, the
parallel run did much less work than the sequential run: a speedup of over 100 was obtained from 20
cores. A similar effect occurred with ``cond-mat'' and $k=4$. This is because the work splitting
mechanism we used explicitly diversifies at the top of search first, where branching heuristics are
least likely to be correct, which leads to an initial incumbent being found faster---this is in line
with recent observations that tailored work stealing should be favoured over randomised work
stealing for combinatorial search problems
\cite{DBLP:conf/cp/ChuSS09,DBLP:journals/topc/McCreeshP15}.

@@

An Erd\H{o}s-R\'{e}nyi random graph $G(n, p)$ has $n$ vertices, and an edge between each distinct
pair of vertices with probability $p$, chosen independently. We now investigate the size of a
maximum $k$-clique in such graphs, and the complexity of finding it. In each case, we use an average
over 100 samples for every point.  We do not use the domination rule for these experiments: the
probability of random graphs having dominating vertices is very low.

@@

We have shown that using a maximum clique algorithm to solve the maximum $k$-clique algorithm for a
graph $G$ by considering $G^k$ in place of $G$ is feasible in practice. This is despite $G^k$
potentially being dense even if $G$ is sparse---this ruled out the use of maximum clique algorithms
which are designed for sparse graphs
\cite{DBLP:conf/waw/PattabiramanPGLC13,DBLP:journals/cor/SegundoLP16}, and we were working with
graphs with many more vertices than is typical for dense maximum clique algorithms.

We introduced a new lazy global domination rule. This was sometimes extremely beneficial, leading to
exponential reductions in the search space---without this rule, we would have been unable to solve
nine of the problem instances we considered, and many others would have taken much longer. However,
even with laziness there is still sometimes a cost to pay when this rule does nothing. This rule is
thus harmful (although only by a polynomial factor) for the graphs typically considered for the
maximum clique problem, and we see the benefit of tailoring algorithms to the problem being solved.
We suggest that a similar rule may also be useful for the maximum $k$-club problem.

We were able to use parallelism to close two further instances, although this required a much finer
level of task granularity than usual.  We suspect further progress could be made by tailoring the
initial vertex ordering based upon what we know about the graphs, or by increasing the number of
recursive calls per second by making the colouring stage cheaper, for example by reusing colour
classes \cite{DBLP:conf/lion/NikolaevBS15}.
